Online Multiclass Linear Classification with Bandit Feedback

### 2019-01-14 ICML-submission/from-linear-separability-to-strong-linear-separability.tex
First pass over section 5

In this section, we show how to construct a mapping $\phi$ from the unit ball of
$\R^d$ into a high dimensional inner product space that has the property that if
a set of labeled examples in the unit ball is linearly separable with a margin
$\gamma$, applying the mapping $\phi$ makes the examples \emph{strongly}
linearly separable with a margin $\gamma'$ and their norms are bounded by $R'$.
The parameters $\gamma'$ and $R'$ are functions of the old margin $\gamma$ and
the number of classes $K$, and are specified in the theorems below.

Equipped with the mapping $\phi$, we can utilize
Algorithm~\ref{algorithm:algorithm-for-strongly-linearly-separable-examples} and
\autoref{theorem:strongly-separable-example-mistake-upper-bound} from the
previous section and we obtain an algorithm for linearly separable examples and
an upper bound on its number of mistakes. As a computational speed up, instead
of working with the mapped examples $\phi(x_1), \phi(x_2), \dots$ explicitly, we
can use the kernelized version of
Algorithm~\ref{algorithm:algorithm-for-strongly-linearly-separable-examples}
that uses a kernel function $k(x,x') = \ip{\phi(x)}{\phi(x')}$.

We construct several different mappings $\phi$. The mappings differ in the
parameters $R'$, $\gamma'$, time complexity of evaluating $\phi(x)$, and time
complexity of evaluating $k(x,x')$. Some of the mappings depend the original
margin $\gamma$. In practice, the margin parameter $\gamma$ is not known, which
makes these mappings impractical. However, one of the mappings we construct
does \emph{not} depend on $\gamma$.

The idea behind all the mappings is polynomial approximation. According to the
well known Stone-Weierstrass theorem (see
e.g.~\citep[Section~10.10]{Davidson-Donsig-2010}), on a compact set,
multivariate polynomials uniformly approximate any continuous function.
Intuitively speaking, we use a multivariate polynomial to approximate, on the
unit ball of $\R^d$, the indicator function corresponding to the intersection of
$m=K-1$ halfspaces. Within margin $\gamma$ along the decision boundary, we allow
the polynomial to attain arbitrary value. The polynomial separates examples in
one class from examples in the other classes. To be able to quantify $R'$,
$\gamma'$ and time complexities of evaluating $\phi(x)$ and $k(x,x')$, we need
to quantify certain parameters of the approximating polynomial. We construct two
different polynomials with different parameters. The parameters are quantified
in
Theorems~\ref{theorem:polynomial-approximation-1}~and~\ref{theorem:polynomial-approximation-2}
stated below.

@@

We construct the three different mappings $\phi_n, \psi_n, \rho$. The first two
are parameterized by a non-negative integer $n$. The domain of $\phi_n$ and
$\psi_n$ is $\R^d$. The domain of $\rho$ is the unit ball of $\R^d$. The
co-domain of $\phi_n$ and $\psi_n$ is $\R^q$ where $q = \sum_{i=0}^n
\binom{i+d-1}{d-1}$ equipped with the standard inner product $\ip{x}{x'}_{\R^q} =
\sum_{i=1}^p x_i x'_i$. The co-domain of $\rho$ is the classical real separable Hilbert space
$\ell_2 = \{ x \in \R^\infty ~:~ \sum_{i=1}^\infty x_i^2 < + \infty \}$ equipped
with the standard inner product $\ip{x}{x'}_{\ell_2} = \sum_{i=1}^\infty x_i x'_i$. We
index the coordinates of the co-domain $\R^q$ by $d$-tuples $(\alpha_1,
\alpha_2, \dots, \alpha_d)$ of non-negative integers such that $\alpha_1 +
\alpha_2 + \dots + \alpha_d \le n$. Similarly, we index the coordinates of
$\ell_2$ by $d$-tuples $(\alpha_1, \alpha_2, \dots, \alpha_d)$ of non-negative
integers. The mappings are defined by
\begin{align*}
& \left(\phi_n(x_1, x_2, \dots, x_d)\right)_{(\alpha_1, \alpha_2, \dots, \alpha_d)}
= x_1^{\alpha_1} x_2^{\alpha_2} \dots x_d^{\alpha_d}
\\
& \left(\psi_n(x_1, x_2, \dots, x_d)\right)_{(\alpha_1, \alpha_2, \dots, \alpha_d)} = x_1^{\alpha_1} x_2^{\alpha_2} \dots x_d^{\alpha_d} \\
& \qquad \cdot \sqrt{\binom{n}{\alpha_1 + \alpha_2 + \dots + \alpha_d} \binom{\alpha_1 + \alpha_2 + \dots + \alpha_d}{\alpha_1, \alpha_2, \dots, \alpha_d}}
\\
& \left(\rho(x_1, x_2, \dots, x_d)\right)_{(\alpha_1, \alpha_2, \dots, \alpha_d)} = x_1^{\alpha_1} x_2^{\alpha_2} \dots x_d^{\alpha_d} \\
& \qquad \cdot \sqrt{2^{-(\alpha_1 + \alpha_2 + \dots + \alpha_d)} \binom{\alpha_1 + \alpha_2 + \dots + \alpha_d}{\alpha_1, \alpha_2, \dots, \alpha_d}}
\end{align*}
where $\binom{\alpha_1 + \alpha_2 + \dots + \alpha_d}{\alpha_1, \alpha_2, \dots, \alpha_d} = \frac{(\alpha_1 + \alpha_2 + \dots + \alpha_d)!}{\alpha_1! \alpha_2! \dots \alpha_d!}$
is the multinomial coefficient and $\binom{n}{k} = \frac{n!}{(n-k)!k!}$ is the binomial coefficient.

Mappings $\psi_n$ and $\rho$ have simple-to-compute kernels
\begin{align*}
K_{\psi_n}(x,x') & = \ip{\psi_n(x)}{\psi_n(x')}_{\R^q} = \left( 1 + \ip{x}{x'} \right)^n \; , \\
K_{\rho}(x,x') & = \ip{\rho(x)}{\rho(x')}_{\ell_2} = \frac{1}{1 - \frac{1}{2}\ip{x}{x'}} \; .
\end{align*}
In passing we note that the formula for $K_\rho$ implies the mapping $\rho$ is
well defined. Indeed, since $K_{\rho}(x,x) < +\infty$ for any $x$ in the unit
ball of $\R^d$, it means that $\rho(x)$ indeed lies in $\ell_2$.

Any mutlivariate polynomial in the orginal space $\R^d$ can be expressed as a
linear function in the new feature spaces given by mappings $\phi_n, \psi_n,
\rho$. The following lemma bounds the norm of the linear function in the terms
of the norm of the polynomial. Recall that a linear function in a Hilbert space
$(H, \ip{\cdot}{\cdot}_H)$ has the form $\ip{c}{x}_H$ for some vector $c \in H$
and its norm is $\norm{c}_H = \sqrt{\ip{c}{c}}$.

\begin{lemma}[Norm bound]
\label{lemma:norm-bound}
Let $d \ge 1$ and $n \ge 0$ be integers and let $q = \sum_{i=0}^n \binom{i+d-1}{d-1}$.
Let $p:\R^d \to \R$ be a multivariate polynomial.
\begin{enumerate}
\item If $n \ge \deg(p)$ then there exists $c \in \R^q$ such that $p(x) = \ip{c}{\phi_n(x)}_{\R^q}$
and $\norm{c}_{\R^q} = \norm{p}$.
\item If $n \ge \deg(p)$ then there exists $c' \in \R^q$ such that $p(x) = \ip{c'}{\psi_n(x)}_{\R^q}$
and $\norm{c'}_{\R^q} \le \norm{p}$.
\item There exists $c'' \in \ell_2$ such that $p(x) = \ip{c''}{\psi_n(x)}_{\ell_2}$
and $\norm{c''}_{\R^q} \le 2^{\deg(p)/2} \norm{p}$.
\end{enumerate}
\end{lemma}

\begin{proof}
Note that the polynomial $p$ can be written as
$p(x) = \sum_{\alpha_1, \alpha_2, \dots, \alpha_d} c_{\alpha_1, \alpha_2, \dots, \alpha_d} x_1^{\alpha_1} x_2^{\alpha_2} \dots x_d^{\alpha_d}$.
There are $\sum_{i=0}^{\deg(p)} \binom{i+d-1}{d-1}$ coefficients $c_{\alpha_1, \alpha_2, \dots, \alpha_d}$.
If $n \ge \deg(p)$ these can be expressed as a vector in $c \in \R^q$. Thus, $p(x) = \ip{c}{\phi_n(x)}_{\R^q}$
and
$$
\norm{c}_{\R^p} = \sqrt{\sum_{\alpha_1, \alpha_2, \dots, \alpha_d} (c_{\alpha_1, \alpha_2, \dots, \alpha_d})^2} =  \norm{p} \; .
$$

We define $c' \in \R^q$ using the multi-index notation as
$$
c'_{\alpha_1, \alpha_2, \dots, \alpha_d}
= \frac{c_{\alpha_1, \alpha_2, \dots, \alpha_d}}{\sqrt{\binom{n}{\alpha_1 + \alpha_2 + \dots + \alpha_d} \binom{\alpha_1 + \alpha_2 + \dots + \alpha_d}{\alpha_1, \alpha_2, \dots, \alpha_d}}} \; .
$$
We have $\ip{c'}{\psi_n(x)}_{\R^q} = \ip{c}{\phi_n(x)}_{\R^q} = p(x)$. Since
$$
|c'_{\alpha_1, \alpha_2, \dots, \alpha_d}| \le |c_{\alpha_1, \alpha_2, \dots, \alpha_d}| \; ,
$$
we have $\norm{c'}_{\R^q} \le \norm{c}_{\R^q} = \norm{p}$.

Similarly, we define $c'' \in \ell_2$ using the multi-index notation as
$$
c''_{\alpha_1, \alpha_2, \dots, \alpha_d}
= \frac{c_{\alpha_1, \alpha_2, \dots, \alpha_d} 2^{(\alpha_1 + \alpha_2 + \dots + \alpha_d)/2}}{\sqrt{\binom{\alpha_1 + \alpha_2 + \dots + \alpha_d}{\alpha_1, \alpha_2, \dots, \alpha_d}}}
$$
for all tuples $(\alpha_1, \alpha_2, \dots, \alpha_d)$ such that $\alpha_1 + \alpha_2 + \dots + \alpha_d \le \deg(p)$.
Otherwise, we define $c''_{\alpha_1, \alpha_2, \dots, \alpha_d} = 0$. We have
$\ip{c''}{\rho(x)}_{\ell_2} = \ip{c}{\phi_n(x)}_{\R^q} = p(x)$. Since
\begin{align*}
|c''_{\alpha_1, \alpha_2, \dots, \alpha_d}|
& \le 2^{(\alpha_1 + \alpha_2 + \dots + \alpha_d)/2} |c''_{\alpha_1, \alpha_2, \dots, \alpha_d}| \\
& \le 2^{\deg(p)/2} |c''_{\alpha_1, \alpha_2, \dots, \alpha_d}| \; ,
\end{align*}
we have $\norm{c''}_{\ell_2} \le 2^{\deg(p)/2} \norm{c}_{\R^q} = 2^{\deg(p)/2} \norm{p}$.
\end{proof}

@@

\begin{theorem}
Let $(x_1, y_1), (x_2, y_2), \dots, (x_T, y_T) \in \R^d \times \{1,2,\dots,K\}$
labeled examples that are weakly linearly separable with margin $\gamma > 0$
such that $\norm{x_1}, \norm{x_2}, \dots, \norm{x_T} \le 1$. Then, the labeled
examples $(\rho(x_1), y_1), (\rho(x_2), y_2), \dots, (\rho(x_T), y_T)$ are
strongly linearly separable with margin
$$
\gamma' = \frac{1}{2\sqrt{K} \left(376 \lceil \log_2(2m) \rceil \cdot \left \lceil \sqrt{\frac{2}{\gamma}} \right \rceil \right)^{\frac{1}{2} \lceil \log_2(2m) \rceil \cdot \left \lceil \sqrt{\frac{2}{\gamma}} \right \rceil}}
$$
and $\norm{\rho(x_1)}_{\ell_2}, \norm{\rho(x_2)}_{\ell_2}, \dots, \norm{\rho(x_T)}_{\ell_2} \le 2$.
TODO
\end{theorem}

>>>>

In this section, we construct and analyze a mapping $\phi$ from the unit ball of
$\R^d$ into a infinite-dimensional inner product space that has the property
that if a set of labeled examples in the unit ball is \emph{weakly} linearly
separable with a margin $\gamma$, applying the mapping $\phi$ makes the examples
\emph{strongly} linearly separable with a margin $\gamma'$ and their norms are
bounded by $2$. The parameter $\gamma'$ is a function of the old margin $\gamma$
and the number of classes $K$, and is specified in the
\autoref{theorem:margin-transformation} below.

The domain of $\phi$ is the unit ball of $\R^d$. The co-domain of $\phi$ is the classical real separable Hilbert space
$\ell_2 = \{ x \in \R^\infty ~:~ \sum_{i=1}^\infty x_i^2 < + \infty \}$ equipped
with the standard inner product $\ip{x}{x'}_{\ell_2} = \sum_{i=1}^\infty x_i x'_i$. We index the coordinates of
$\ell_2$ by $d$-tuples $(\alpha_1, \alpha_2, \dots, \alpha_d)$ of non-negative
integers. The mapping is defined by
\begin{align*}
& \left(\phi(x_1, x_2, \dots, x_d)\right)_{(\alpha_1, \alpha_2, \dots, \alpha_d)} = x_1^{\alpha_1} x_2^{\alpha_2} \dots x_d^{\alpha_d} \\
& \qquad \cdot \sqrt{2^{-(\alpha_1 + \alpha_2 + \dots + \alpha_d)} \binom{\alpha_1 + \alpha_2 + \dots + \alpha_d}{\alpha_1, \alpha_2, \dots, \alpha_d}}
\end{align*}
where $\binom{\alpha_1 + \alpha_2 + \dots + \alpha_d}{\alpha_1, \alpha_2, \dots, \alpha_d} = \frac{(\alpha_1 + \alpha_2 + \dots + \alpha_d)!}{\alpha_1! \alpha_2! \dots \alpha_d!}$
is the multinomial coefficient. The corresponding kernel is
$$
k(x,x') = \ip{\phi(x)}{\phi(x')}_{\ell_2} = \frac{1}{1 - \frac{1}{2}\ip{x}{x'}} \; .
$$
In passing we note that the formula for $k$ implies the mapping $\phi$ is
well defined. Indeed, since $k(x,x) < +\infty$ for any $x$ in the unit
ball of $\R^d$, it means that $\phi(x)$ indeed lies in $\ell_2$.

Equipped with the mapping $\phi$, we can utilize
Algorithm~\ref{algorithm:kernelized} with kernel function $k(x,x') =
\ip{\phi(x)}{\phi(x')}$ and
\autoref{theorem:strongly-separable-example-mistake-upper-bound} from the
previous section and we obtain an algorithm for weakly linearly separable
examples and an upper bound on its number of mistakes.

The idea behind construction and analysis of the mapping $\phi$ is polynomial
approximation. According to the well known Stone-Weierstrass theorem (see
e.g.~\citet[Section~10.10]{Davidson-Donsig-2010}), on a compact set,
multivariate polynomials uniformly approximate any continuous function.
Intuitively speaking, we use a multivariate polynomial to approximate, on the
unit ball of $\R^d$, the indicator function corresponding to the intersection of
$m=K-1$ halfspaces. Within margin $\gamma$ along the decision boundary, we allow
the polynomial to attain arbitrary value. The polynomial separates examples in
one class from examples in the other classes. To be able to quantify $R'$,
$\gamma'$ and time complexities of evaluating $\phi(x)$ and $k(x,x')$, we need
to quantify certain parameters of the approximating polynomial. We construct two
different polynomials with different parameters. The parameters are quantified
in
Theorems~\ref{theorem:polynomial-approximation-1}~and~\ref{theorem:polynomial-approximation-2}
stated below.

@@

The following lemma expresses any multivariate polynomial in $\R^d$
as a linear function in $\ell_2$ and gives an upper bound its the norm.

\begin{lemma}[Norm bound]
\label{lemma:norm-bound}
Let $p:\R^d \to \R$ be a multivariate polynomial.
There exists $c \in \ell_2$ such that $p(x) = \ip{c}{\psi_n(x)}_{\ell_2}$
and $\norm{c}_{\ell_2} \le 2^{\deg(p)/2} \norm{p}$.
\end{lemma}

\begin{proof}
Note that the polynomial $p$ can be written as
$p(x) = \sum_{\alpha_1, \alpha_2, \dots, \alpha_d} c'_{\alpha_1, \alpha_2, \dots, \alpha_d} x_1^{\alpha_1} x_2^{\alpha_2} \dots x_d^{\alpha_d}$.
We define $c \in \ell_2$ using the multi-index notation as
$$
c_{\alpha_1, \alpha_2, \dots, \alpha_d}
= \frac{c'_{\alpha_1, \alpha_2, \dots, \alpha_d} 2^{(\alpha_1 + \alpha_2 + \dots + \alpha_d)/2}}{\sqrt{\binom{\alpha_1 + \alpha_2 + \dots + \alpha_d}{\alpha_1, \alpha_2, \dots, \alpha_d}}}
$$
for all tuples $(\alpha_1, \alpha_2, \dots, \alpha_d)$ such that $\alpha_1 + \alpha_2 + \dots + \alpha_d \le \deg(p)$.
Otherwise, we define $c_{\alpha_1, \alpha_2, \dots, \alpha_d} = 0$. Clearly,
$\ip{c}{\phi(x)}_{\ell_2} = p(x)$.

Since
\begin{align*}
|c_{\alpha_1, \alpha_2, \dots, \alpha_d}|
& \le 2^{(\alpha_1 + \alpha_2 + \dots + \alpha_d)/2} |c'_{\alpha_1, \alpha_2, \dots, \alpha_d}| \\
& \le 2^{\deg(p)/2} |c'_{\alpha_1, \alpha_2, \dots, \alpha_d}| \; ,
\end{align*}
we have
\begin{align*}
\norm{c}_{\ell_2}
& \le 2^{\deg(p)/2} \sqrt{\sum_{\alpha_1, \alpha_2, \dots, \alpha_d} (c'_{\alpha_1, \alpha_2, \dots, \alpha_d})^2} \\
& = 2^{\deg(p)/2} \norm{p} \; .
\end{align*}
\end{proof}

@@

\begin{theorem}[Margin transformation]
\label{theorem:margin-transformation}
Let $(x_1, y_1), (x_2, y_2), \dots, (x_T, y_T) \in \R^d \times \{1,2,\dots,K\}$
labeled examples that are weakly linearly separable with margin $\gamma > 0$
such that $\norm{x_1}, \norm{x_2}, \dots, \norm{x_T} \le 1$. Then, the labeled
examples $(\rho(x_1), y_1), (\rho(x_2), y_2), \dots, (\rho(x_T), y_T)$ are
strongly linearly separable with margin
$$
\gamma' = \frac{1}{2\sqrt{K} \left(376 \lceil \log_2(2m) \rceil \cdot \left \lceil \sqrt{\frac{2}{\gamma}} \right \rceil \right)^{\frac{1}{2} \lceil \log_2(2m) \rceil \cdot \left \lceil \sqrt{\frac{2}{\gamma}} \right \rceil}}
$$
and $\norm{\rho(x_1)}_{\ell_2}, \norm{\rho(x_2)}_{\ell_2}, \dots, \norm{\rho(x_T)}_{\ell_2} \le 2$.
TODO
\end{theorem}
